Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2020-05-15 Manuscript/manuscript.Rmd
editted discussion, first 2 paragraphs

In this study we developed an ensemble framework to systematically compare the effects of different combinations of Hi-C resolution, data re-sampling, and feature engineering techniques on the performance of models aimed at TAD boundary prediction. By implementing our framework across multiple cell lines, we showed that there exists a consensus approach toward precise TAD boundary prediction. We showed that random forest classification models trained on domain data extracted from 10 kb resolution maps, using random over- and under-sampling to account for class imbalance, with a distance-type predictor space yielded optimal model performances for both the GM12878 and K562 cell lines.

Previous studies tended to focus only on specific sets of genomic annotations including chromatin states, histone modifications, or transcription factor binding sites (TFBS) alone. Using our framework, we were able to conclude that simply building random forest classifiers only on cell line specific TFBS can yield near optimal model performance, thereby significantly reducing the feature space and computational complexity of the model. Specifically, we found that the CCCTC-binding factor (CTCF) and cohesin complexes (RAD21 and SMC3) were the most important features of TAD boundary prediction, confirming their enrichment at TAD boundaries as part of the loop extrusion model [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Likewise, we found the the zinc-finger protein ZNF143 was also given high predictive importance. This confirmed the known role of ZNF143 as a chromatin-looping factor that contributes to the structural foundation of the genome by providing sequence specificity at promoters connected with distal regulatory elements [@bailey2015znf143;@gonzalez2017znf143].

By leveraging the enrichment of these architectural proteins we were able to more precisely identify TAD boundaries at a much more refined resolution. To do so, we evaluated the ChIP-Seq signal strength between the CTCF, RAD21, SMC3, and ZNF143 transcription factors and a 10 kb flanked window of predicted TAD boundaries identified by our optimally built random forest model. We showed that the genomic signals for each TFBS was much more consolidated near the center of our flanked predicted TAD boundary then the flanked boundary identified by the ARROWHEAD TAD caller. We also evaluated the distal relationships between the same architectural proteins and found that they were statistically significantly closer in base pair distance to our predicted TAD boundaries, than to those identified by ARROWHEAD.

>>>>

Here we present *preciseTAD*, a random forest-based approach coupled with density-based spatial clustering (DBSCAN) and scalable partition around medoids (PAM/CLARA) to more precisely predict TAD or loop boundaries on the linear genome. *preciseTAD* leverages a random forest (RF) classification algorithm built on ground truth TAD(loop) boundaries obtained from a user defined TAD(loop) calling tool and ChIP-seq defined genomic annotations as the feature space. We optimized our RF model by creating a computational framework that systematically compared the performances of different combinations of genome binning (resolution), feature engineering procedure, and re-sampling technique. In doing so, we first introduced a new strategy for establishing boundary prediction data referred to as *shifted binning*, in which the rows of the data matrix were controlled by the data resolution (either 5kb, 10kb, 25kb, 50kb, and 100kb) used to obtain ground truth boundaries, while assuring ground truth bounderies were located central to any given genomic bin. Furthermore, we introduced a new procedure for drawing spatial associations between boundary regions and genomic annotations, referred to as *distance*-type predictors. The method was benchmarked against two boundary calling tools, ARROWHEAD [@durand2016juicebox], an established TAD-caller, and PEAKACHU [@salameh2019supervised] a recently published algorithm for predicting chromatin loops. For a more in depth study, we applied *preciseTAD* on cell line-specific boundaries and ChIP-seq data from the GM12878 and K562 cell lines.

Our computational framework alone yielded many interesting results. First, we showed that RF models built using *distance*-type predictors out performed models built on previously published feature engineering techniques including overlap counts and overlap percents. Moreover, we were further able to optimize our RF model by introducing random under-sampling, a technique to correct for the severe class imbalance between TAD(loop) and non-TAD(loop) boundary regions. We showed that in our optimized RF, transcription factor binding sites provided an overwhelming amount of signal in the model, indeed out-performing histone modification- and BroadHMM-specific models, contradicting a previous study that found histone marks alone were effective in predicting TAD boundaries [@huang2015predicting]. Likewise, our RF model built using *distance*-type predictors was consistently able to identify the group of the top known molecular drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143 as being highly predictive of TAD(loop) boundary regions across both cell lines. These are known components of the loop extrusion model, an established theory for how loops are made by a ring-shaped adenosine triphosphatase complex called cohesin [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. This suggested that random forests, when tuned and feature engineered correctly, are highly effective in accurately predicting boundary regions, contradicting a previous study that found simple multiple linear regression out performed random forests in a TAD boundary prediction setting using both human and drosophila Hi-C data [@mourad2016computational]. Plausible reasons for their findings include lack of class balancing prior to modeling and usage of rudimentary feature engineering procedures including overlap counts and overlap percents.

???stopped here
